---
title: "Exploration and Summary"
author: "Mugeng Liu"
date: "2024-09-30"
output:
  pdf_document: default
  html_document: default
---
```{r}
install.packages("ggplot2", repos = "https://cloud.r-project.org/")
```

```{r}
library(here)
library(ggplot2)
finaldata <- read.csv(here("data/analytical", "finaldata.csv"), header = TRUE)
names(finaldata)
```

First, look at the summary of variables.

```{r}
finaldata |>
  summary()
```

General Statistics: The observations are from 186 countries and across 20 years, which means total of 3720 data. We see a very large right skewed distribution in the variable gdp1000, totdeat, matmor, infmor, un5mor. Only temp, male_edu is left skewed. This shows that countries not only have large variation, but also few countries have better condition. The Max in gdp1000 is 123,6787 whereas median is 4.0719 and mean is 11.4917.

```{r}
rows_u <- finaldata[is.na(finaldata$matmor), c("country_name", "year")]

length(unique(rows_u$country_name))
print("number of country missed in matmor")

rows_um <- finaldata[is.na(finaldata$neomor), c("country_name", "year")]
rows_um
print("number of country missed in matmor")

```

Missing Data: We then check the missing data in our observation. Variable Matmor has the biggest portion of data missing, about 426 in 3720 are missed, account for 186 countries in different year, which means all countries have data missing in different year. All countries missed data in 2018 and 2019, suggested a MCAR type missing data due to unable to update the data. It is noticeable that infmor, neomor, un5mor variables missed 20 observations. It is the country Puerto Rico missed these observations across the whole interval from 2000 to 2019.

Relationship between matmor, infmor, neomor and un5mor. We also want to explore whether there's association between these variables.

```{r}
finaldata |>
  ggplot(aes(x = matmor, y = infmor, alpha = 0.1)) +
  geom_point()

finaldata |>
  ggplot(aes(x = matmor, y = infmor, alpha = 0.1)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10()


```

```{r}
finaldata |>
  ggplot(aes(x = infmor, y = neomor, alpha = 0.1)) +
  geom_point()

finaldata |>
  ggplot(aes(x = infmor, y = neomor, alpha = 0.1)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10()
```

```{r}
finaldata |>
  ggplot(aes(x = neomor, y = un5mor, alpha = 0.1)) +
  geom_point()

finaldata |>
  ggplot(aes(x = neomor, y = un5mor, alpha = 0.1)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10()
```

Relationship between gdp1000 and other variables

```{r}
variables <- setdiff(names(finaldata), "gdp1000")

# Create 23 ggplot graphs
for (var in variables) {
  p <- ggplot(finaldata, aes_string(x = "gdp1000", y = var)) +
    geom_point(alpha = 0.5) +
    labs(title = paste("gdp1000 vs", var), x = "gdp1000", y = var)

  # Print the plot
  print(p)
}
```

Next, apply the log transformation for those continous variables

```{r}
variables <- setdiff(names(finaldata), "gdp1000")

# Create 23 ggplot graphs with log scales
variables <- setdiff(names(finaldata), "gdp1000")
numeric_variables <- variables[sapply(finaldata[variables], is.numeric)]  # Filter numeric variables only

# Create ggplot graphs with log scales for numeric variables
for (var in numeric_variables) {
  p <- ggplot(finaldata, aes_string(x = "gdp1000", y = var)) +
    geom_point(alpha = 0.5) +
    scale_x_log10() +  # Apply log scale to the x-axis
    scale_y_log10() +  # Apply log scale to the y-axis
    labs(title = paste("gdp1000 (log scale) vs", var), x = "gdp1000 (log scale)", y = paste(var, "(log scale)"))

  # Print the plot
  print(p)
}
```

We then compute the coefficient of determination to see the relation between gdp1000 and other variables

```{r}
variables <- setdiff(names(finaldata), "gdp1000")
numeric_variables <- variables[sapply(finaldata[variables], is.numeric)]  # Filter numeric variables only

# Calculate and print the coefficient of determination (R²)
for (var in numeric_variables) {
  correlation <- cor(finaldata$gdp1000, finaldata[[var]], use = "complete.obs")  # Calculate correlation
  r_squared <- correlation^2  # Coefficient of determination (R²)
  print(paste("R² between gdp1000 and", var, ":", r_squared))  # Print R²
}
```

Since there are constant for some vairables, we use log transformation to further explore the relationship

```{r}
# Get the variable names excluding 'gdp1000' and keeping only numeric variables
variables <- setdiff(names(finaldata), "gdp1000")
numeric_variables <- variables[sapply(finaldata[variables], is.numeric)]  # Filter numeric variables only

# Calculate and print the coefficient of determination (R²) using log scale
for (var in numeric_variables) {
  # Take the log of both gdp1000 and the variable, and handle non-positive values by removing them
  log_gdp1000 <- log10(finaldata$gdp1000[finaldata$gdp1000 > 0 & finaldata[[var]] > 0])
  log_var <- log10(finaldata[[var]][finaldata$gdp1000 > 0 & finaldata[[var]] > 0])
  
  # Calculate correlation on the log-transformed values
  correlation <- cor(log_gdp1000, log_var, use = "complete.obs")  
  r_squared <- correlation^2  # Coefficient of determination (R²)
  
  # Print the result
  print(paste("R² (log scale) between gdp1000 and", var, ":", r_squared))
}
```

Summary of relation between variables: 1. There are very strong positive correlations on variables between matmor, infmor, neomor, un5mor. 2. There are very strong correlations on variable gdp1000 with variables agedep, matmor, infmor, neomor, un5mor. Un5mor has the highest correlation of determination, which accounts for 0.76 in log transformation. In addition, it shows that gdp1000 and mo rate has very high negative correlation, gdp1000 also have negative correlation with variable agedep and positive correlation with male_edu.
